\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-11 papers/jss/article.Rnw
more edits
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@751 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

One key extension of \pkg{RProtoBufUtils} is the 
\texttt{serialize\_pb} method to convert R objects into serialized
Protocol Buffers in the catch-all schema. The \texttt{can\_serialize\_pb}
method can then be used to determine whether the given R object can safely
be expressed in this way.  To illustrate how this method works, we
attempt to convert all of the built-in datasets from R into this
serialized Protocol Buffer representation.

@@

<<echo=TRUE>>=
#datasets$valid.proto <- sapply(datasets$load.name, function(x) can_serialize_pb(eval(as.name(x))))
#datasets <- subset(datasets, valid.proto==TRUE)
m <- nrow(datasets)
@

\Sexpr{m} data sets could be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%).  The next section illustrates how
many bytes were used to store the data sets under four different
situations (1) normal R serialization, (2) R serialization followed by
gzip, (3) normal Protocol Buffer serialization, (4) Protocol Buffer
serialization followed by gzip.

@@

Table~\ref{tab:compression} shows the sizes of 50 sample R datasets as
returned by object.size() compared to the serialized sizes.
The summary compression sizes are listed below, and a full table for a
sample of 50 datasets is included on the next page.  Sizes are comparable
but protocol buffers provide simple getters and setters in multiple
languages instead of requiring other programs to parse the R
serialization format \citep{serialization}.  One takeaway from this
table is that RProtoBuf does not in general provide any significant
space-savings over R's normal serialization mechanism.  The benefit
from RProtoBuf comes from its interoperability with other
environments, safe versioning,

TODO comparison of protobuf serialization sizes/times for various vectors.  Compared to R's native serialization.  Discussion of the RHIPE approach of serializing any/all R objects, vs more specific protocol buffers for specific R objects.

@@

TODO RProtoBuf is quite flexible and easy to use for interactive
analysis, but it is not designed for certain classes of operations one
might like to do with protocol buffers.  For example, taking a list of
10,000 protocol buffers, extracting a named field from each one, and
computing a aggregate statistics on those values would be extremely
slow with RProtoBuf, and while this is a useful class of operations,
it is outside of the scope of RProtoBuf.  We should be very clear
about this to clarify the goals and strengths of RProtoBuf and its
reflection and object mapping.

@@

As described earlier, the primary application of protocol buffers is
data interchange in the context of inter-system communications. 
Network protocols such as HTTP provide mechanisms for client-server
communication, i.e. how to initiate requests, authenticate, send messages, 
etc.  However, many network 
protocols generally do not regulate \emph{content} of messages: they allow
transfer of any media type, such as web pages, files or video.
When designing systems where various components require exchange of specific data
structures, we need something on top of the network protocol that prescribes 
how these structures are to be represented in messages (buffers) on the
network. Protocol buffers solve exactly this problem by providing
a cross platform method for serializing arbitrary structures into well defined
messages, that can be exchanged using any protocol. The descriptors
(\texttt{.proto} files) are used to formally define the interface of a
remote API or network application. Libraries to parse and generate protobuf
messages are available for many programming languages, making it 
relatively straight forward to implement clients and servers.

@@

One example of a system that supports protocol buffers to interact
with R is OpenCPU \citep{opencpu}. OpenCPU is a framework for embedded statistical 
computation and reproducible research based on R and \LaTeX. It exposes a 
HTTP(S) API to access and manipulate R objects and allows for performing 
remote R function calls. Clients do not need to understand 
or generate any R code: HTTP requests are automatically mapped to 
function calls, and arguments/return values can be posted/retrieved
using several data interchange formats, such as protocol buffers.  
OpenCPU uses the \texttt{serialize\_pb} and \texttt{unserialize\_pb} functions
from the \texttt{RProtoBuf} package to convert between R objects and protobuf
messages. Therefore, clients need the \texttt{rexp.proto} descriptor mentioned
earlier to parse and generate protobuf messages when interacting with OpenCPU.

@@

# Retrieve and parse message
req <- GET ('https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb')
output <- unserialize_pb(req$content)

@@

This code suggests a method for exchanging objects between R servers, however this can 
also be done without protocol buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

\begin{verbatim}
  protoc rexp.proto --python_out=.
\end{verbatim}
This generates python module called \texttt{rexp\_pb2.py}, containing both the 
descriptor information as well as methods to read and manipulate the R object 
message. In the example below we use the HTTP client from the \texttt{urllib2}
module.

@@

The example above shows how the \texttt{HTTP GET} method retrieves a 
resource from OpenCPU, for example an R object. The \texttt{HTTP POST} 
method on the other hand is used for calling functions and running scripts, 
which is the primary purpose of the framework. As before, the \texttt{/pb} 
postfix requests to retrieve the output as a protobuf message, in this
case the function return value. However, OpenCPU allows us to supply the
arguments of the function call in the form of protobuf messages as well.
This is a bit more work, because clients needs to both generate messages 
containing R objects to post to the server, as well as retrieve and parse
protobuf messages returned by the server. Using protocol buffers to post
function arguments is not required, and for simple (scalar) arguments 
the standard \texttt{application/x-www-form-urlencoded} format might be sufficient.
However, with protocol buffers the client can perform function calls with
more complex arguments such as R vectors or lists. The result is a complete
RPC system to do arbitrary R function calls from within 
any programming language.

@@

<<eval=FALSE>>=
fnargs <- unserialize_pb(inputmsg)
val <- do.call(stats::rnorm, fnargs)
outputmsg <- serialize_pb(val)
@
In reality the OpenCPU provides a lot of meta functionality such as handling
of sessions, exceptions, security, and much more. OpenCPU also makes it possible to store
output of a function call on the server, instead of directly retrieving it. Thereby 
objects can be shared with other users or used as arguments in a subsequent
function call. But in its essence, the HTTP API provides a simple way to perform remote 
R function calls over HTTPS. The same request can be performed in Python as follows:

@@

The first versions of \CRANpkg{RProtoBuf} were written during 2009-2010,
with very significant contributions, both in code and design, made by
Romain Fran\c{c}ois. His continued influence on design and code is
appreciated. Several features of the package are influenced
by the design of the \CRANpkg{rJava} package by Simon Urbanek
The user-defined table mechanism, implemented by Duncan Temple Lang for the
purpose of the \pkg{RObjectTables} package allowed the dynamic symbol lookup.
Kenton Varda was generous with his time in reviewing code and explaining
obscure protocol buffer semantics.  Karl Millar and Jeroen Ooms were
helpful in reviewing code or offering suggestions.  The contemporaneous
work by Saptarshi Guha on \pkg{RHIPE} was a strong initial motivator.

>>>>

One key extension which \pkg{RProtoBufUtils} brought to \pkg{RProtoBuf} is the 
\texttt{serialize\_pb} method to convert R objects into serialized
Protocol Buffers in the catch-all schema. The \texttt{can\_serialize\_pb}
method can then be used to determine whether the given R object can safely
be expressed in this way.  To illustrate how this method works, we
attempt to convert all of the built-in datasets from R into this
serialized Protocol Buffer representation.

@@

<<echo=TRUE>>=
datasets$valid.proto <- sapply(datasets$load.name,
                        function(x) can_serialize_pb(eval(as.name(x))))
datasets <- subset(datasets, valid.proto==TRUE)
m <- nrow(datasets)
@

\Sexpr{m} data sets could be converted to Protocol Buffers
(\Sexpr{format(100*m/n,digits=1)}\%).  The next section illustrates how
many bytes were used to store the data sets under four different
situations:
\begin{itemize}
\item normal R serialization,
\item R serialization followed by gzip,
\item normal Protocol Buffer serialization, and
\item Protocol Buffer serialization followed by gzip.
\end{itemize}

@@

Table~\ref{tab:compression} shows the sizes of 50 sample R datasets as
returned by object.size() compared to the serialized sizes.
%The summary compression sizes are listed below, and a full table for a
%sample of 50 datasets is included on the next page.  
Sizes are comparable but Protocol Buffers provide simple getters and setters
in multiple languages instead of requiring other programs to parse the R
serialization format.% \citep{serialization}.
One takeaway from this table is that RProtoBuf does not in general provide
any significant saving in file size compared to the normal serialization
mechanism in R which is seen as equally compact.  The benefit from RProtoBuf
comes from its interoperability with other environments, as well as its safe
versioning,

TODO comparison of protobuf serialization sizes/times for various vectors.
Compared to R's native serialization.  Discussion of the RHIPE approach of
serializing any/all R objects, vs more specific Protocol Buffers for specific
R objects.

@@

TODO RProtoBuf is quite flexible and easy to use for interactive
analysis, but it is not designed for certain classes of operations one
might like to do with Protocol Buffers.  For example, taking a list of
10,000 Protocol Buffers, extracting a named field from each one, and
computing a aggregate statistics on those values would be extremely
slow with RProtoBuf, and while this is a useful class of operations,
it is outside of the scope of RProtoBuf.  We should be very clear
about this to clarify the goals and strengths of RProtoBuf and its
reflection and object mapping.

@@

As described earlier, the primary application of Protocol Buffers is data
interchange in the context of inter-system communications.  Network protocols
such as HTTP provide mechanisms for client-server communication, i.e. how to
initiate requests, authenticate, send messages, etc.  However, many network
protocols generally do not regulate the \emph{content} of messages: they
allow transfer of any media type, such as web pages, static files or
multimedia content.  When designing systems where various components require
exchange of specific data structures, we need something on top of the network
protocol that prescribes how these structures are to be represented in
messages (buffers) on the network. Protocol Buffers solve exactly this
problem by providing a cross-platform method for serializing arbitrary
structures into well defined messages, which can then be exchanged using any
protocol. The descriptors (\texttt{.proto} files) are used to formally define
the interface of a remote API or network application. Libraries to parse and
generate protobuf messages are available for many programming languages,
making it relatively straightforward to implement clients and servers.

@@

One example of a system that supports Protocol Buffers to interact
with R is OpenCPU \citep{opencpu}. OpenCPU is a framework for embedded statistical 
computation and reproducible research based on R and \LaTeX. It exposes a 
HTTP(S) API to access and manipulate R objects and allows for performing 
remote R function calls. Clients do not need to understand 
or generate any R code: HTTP requests are automatically mapped to 
function calls, and arguments/return values can be posted/retrieved
using several data interchange formats, such as protocol buffers.  
OpenCPU uses the \texttt{serialize\_pb} and \texttt{unserialize\_pb} functions
from the \texttt{RProtoBuf} package to convert between R objects and protobuf
messages. Therefore, clients need the \texttt{rexp.proto} descriptor mentioned
earlier to parse and generate protobuf messages when interacting with OpenCPU.

@@

# Retrieve and parse message
req <- GET('https://public.opencpu.org/ocpu/library/MASS/data/Animals/pb')
output <- unserialize_pb(req$content)

@@

This code suggests a method for exchanging objects between R servers, however this can 
also be done without Protocol Buffers. The main advantage of using an inter-operable format 
is that we can actually access R objects from within another
programming language. For example, in a very similar fashion we can retrieve the same
dataset in a Python client. To parse messages in Python, we first compile the 
\texttt{rexp.proto} descriptor into a python module using the \texttt{protoc} compiler:

\begin{verbatim}
  protoc rexp.proto --python_out=.
\end{verbatim}
This generates Python module called \texttt{rexp\_pb2.py}, containing both the 
descriptor information as well as methods to read and manipulate the R object 
message. In the example below we use the HTTP client from the \texttt{urllib2}
module.

@@

The example above shows how the \texttt{HTTP GET} method retrieves a 
resource from OpenCPU, for example an R object. The \texttt{HTTP POST} 
method on the other hand is used for calling functions and running scripts, 
which is the primary purpose of the framework. As before, the \texttt{/pb} 
postfix requests to retrieve the output as a protobuf message, in this
case the function return value. However, OpenCPU allows us to supply the
arguments of the function call in the form of protobuf messages as well.
This is a bit more work, because clients needs to both generate messages 
containing R objects to post to the server, as well as retrieve and parse
protobuf messages returned by the server. Using Protocol Buffers to post
function arguments is not required, and for simple (scalar) arguments 
the standard \texttt{application/x-www-form-urlencoded} format might be sufficient.
However, with protocol buffers the client can perform function calls with
more complex arguments such as R vectors or lists. The result is a complete
RPC system to do arbitrary R function calls from within 
any programming language.

@@

<<eval=FALSE>>=
fnargs <- unserialize_pb(inputmsg)
val <- do.call(stats::rnorm, fnargs)
outputmsg <- serialize_pb(val)
@

OpenCPU also provides a lot of meta-functionality such as handling
of sessions, exceptions, security, and more. OpenCPU also makes it possible to store
output of a function call on the server, instead of directly retrieving it. Thereby 
objects can be shared with other users or used as arguments in a subsequent
function call. But in its essence, the HTTP API provides a simple way to perform remote 
R function calls over HTTPS. The same request can be performed in Python as follows:

@@

The first versions of \CRANpkg{RProtoBuf} were written during 2009-2010.
Very significant contributions, both in code and design, were made by
Romain Fran\c{c}ois whose continued influence on design and code is
greatly appreciated. Several features of the package are influenced
by the design of the \CRANpkg{rJava} package by Simon Urbanek
The user-defined table mechanism, implemented by Duncan Temple Lang for the
purpose of the \pkg{RObjectTables} package, allows for the dynamic symbol lookup.
Kenton Varda was generous with his time in reviewing code and explaining
obscure protocol buffer semantics.  Karl Millar was very
helpful in reviewing code and offering suggestions.  
%The contemporaneous work by Saptarshi Guha on \pkg{RHIPE} was a strong
%initial motivator.
